---
title: "Session 6"
author: "Pia Baader"
date: "2023-06-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

osmdata: street maps 


```{r}
#install.packages("osmdata")
library(osmdata) 
bb <- getbb("Germany")
bb
```
different output formats via format_out

```{r}

```

boundaries plotting: featuretype = ...
tm_shape(bb_test$multipolygon) für das richtige Ergebnis 
constructing an overpass query: 
bb_obb <- getbb("Oberbayern")
q <- opq(bb_obb) %>% add_osm_feature(key= 'natural', value = 'peak')
result <- q %>% osmdata_sf() 
results$osm_points (Suche verfeinern)
etwas auf die Karte: mit tm_shape(result$osm_points) + tm_dots(id="name")


title: "Practical 6"
description: "Fortgeschrittene Statistische Software für NF - SS 2022/23"
author:
  - name: Jan Simson
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Lecture

## Spatial Data

- `{sf}`
- Many formats
- [Shapefiles](https://en.wikipedia.org/wiki/Shapefile) (info via Wikipedia)
  - `.shp` — shape format; the feature geometry itself {content-type: x-gis/x-shapefile}
  - `.shx` — shape index format; a positional index of the feature geometry to allow seeking forwards and backwards quickly {content-type: x-gis/x-shapefile}
  - `.dbf` — attribute format; columnar attributes for each shape, in dBase IV format {content-type: application/octet-stream OR text/plain}
  - `.prj` — projection description, using a well-known text representation of coordinate reference systems {content-type: text/plain OR application/text}

```{r}
library(sf)
library(spData)

bavaria <- read_sf("data/bavaria.shp")
# Nicer column names
colnames(bavaria) <- c(
  "place", "type", "gdp_per_capita", "mean_age",
  "pop_density", "unemployment_rate",
  "employment_rate", "household_income",
  "students", "pop_development", "geometry"
)

bavaria
```

## tmap

```{r}
library(tmap)

tm_shape(bavaria) +
  tm_fill()
```

```{r}
tm_shape(bavaria) +
  tm_borders()
```

### tmap_mode

- `plot` or `view`

```{r}
tmap_mode("plot")

tm_shape(bavaria) +
  tm_borders()
```

```{r}
tmap_mode("view")

tm_shape(bavaria) +
  tm_borders()
```

## Choropleths

```{r}
tmap_mode("plot")

tm_shape(bavaria) +
  tm_polygons(col = "gdp_per_capita")
```

```{r}
tm_shape(bavaria) +
  tm_polygons(col = "pop_density", midpoint = 0)
```

Source of data and example code: https://bookdown.org/nicohahn/making_maps_with_r5/docs/tmap.html

## osmdata

```{r}
library(osmdata)

bb <- getbb("Germany")
bb
```

### Different output formats via `format_out`.

```{r}
bb <- getbb("Germany", format_out = "sf_polygon")
bb
```

Now we can pass it into tmap.

```{r}
tmap_mode("plot")

tm_shape(bb) +
  tm_borders()
```

```{r}
tmap_mode("view")

tm_shape(bb) +
  tm_borders()
```

### Getting bounds for the correct element via `featuretype`

By default getbb will return bounds of type `"settlement"`, anything smaller than a country.

```{r}
bb <- getbb("Germany", featuretype = "country", format_out = "sf_polygon")
bb
```

```{r error=TRUE}
tm_shape(bb) +
  tm_borders()
```

```{r}
tm_shape(bb$multipolygon) +
  tm_borders()
```
```{r}
getbb("Germany", featuretype = "country", format_out = "matrix")
```


### Querying data with `osmdata`

#### Getting the correct bounding box

```{r}
bb_test <- getbb("Munich", format_out = "sf_polygon")
bb_test
```

```{r}
tm_shape(bb_test$multipolygon) +
  tm_borders()
```

### Constructing an overpass Query

For overpapass we need bounding boxes in a different format.

```{r}
bb_obb <- getbb("Oberbayern")
```

https://wiki.openstreetmap.org/wiki/Tag:natural%3Dpeak

```{r}
q <- opq(bb_obb) %>% 
    add_osm_feature(key = 'natural', value = 'peak')
```

The result comes with different kinds of returned data:

```{r}
result <- q %>% 
  osmdata_sf()
result
```

```{r}
result$osm_points
```

```{r}
library(tidyverse)
result$osm_points <- result$osm_points %>% 
  mutate(ele = as.numeric(ele))
```


```{r}
tm_shape(result$osm_points) +
  tm_dots(id = "name")
```
```{r}
library(tidyverse)

result$osm_points %>% 
  mutate(
    ele = as.numeric(ele)
  ) %>% 
  filter(
    ele > 1000
  ) %>% 
  tm_shape() +
  tm_dots(id = "name", col = "ele")
```


# Practical: Let's plan a hiking trip around Munich!

We want to plan a hiking trip in the Munich area and want to use geospatial data science tools to help us with the planning.

## Part 1: Snacks

Before going on a hike we'll need to buy some snacks to eat while up on the peak.

a) Let's start by first getting the bounding box for the `Maxvorstadt` so we can buy some snacks closeby.

```{r}
library(tidyverse)
#install.packages("tmap")
library(osmdata)
library(tmap)

getbb("Maxvorstadt")
```

b) Plot the bounding box using `{tmap}` to check whether it is correct.

```{r}
bb_max <- getbb("Maxvorstadt", format_out = "sf_polygon")

tmap_mode("view")
tm_shape(bb_max) + 
  tm_borders()
```

```

c) Search online to determine the feature type that we will need to find *supermarkets* on OSM. What is the feature type called in the OSM wiki?

shop = supermarket

d) Construct an OSM query to find all supermarkets within the Maxvorstadt.

```{r}
maxv_supermarkets <- opq(getbb("Maxvorstadt")) %>%
  add_osm_feature(key= "shop",  value = "supermarket") %>% 
  osmdata_sf()
  
maxv_supermarkets

maxv_supermarkets$osm_points()
```

e) Create an interactive map of them using tmap.

```{r}
tmap_mode("view")
tm_shape(maxv_supermarkets$osm_points) + 
  tm_dots(id= "name", col= "brand")
```

## Part 2: Destination

a) Let's continue by getting the bounding box for the city of Munich.

```{r}
bb <- getbb("Munich",  format_out = "sf_polygon") %>% 
bb 


```

b) Plot the bounding box using `{tmap}` to check whether it is correct.

```{r}
bb %>% 
  tm_shape(") + 
  tm_borders()
```

c) Construct a query to find all peaks within the Munich area and execute it to get the data.

```{r}
munich_peaks <- opq(getbb("Munich")) %>% 
        add_osm_feature(key="natural", value = "peak") %>% 
        osmdata_sf()
```

d) How many peaks are there in around Munich and what is their average height?

```{r}
munich_peaks$osm_points %>% nrow()

munich_peaks_num <- munich_peaks$osm_points %>% 
mutate(ele= as.numeric(ele))
```

e) Filter the result to only include peaks higher than the median.

```{r}
mp_above_median <- munich_peaks_num %>% 
   filter(ele > median(ele, na.rm = TRUE))
   
mp_above_median
```

f) Use `tm_bubbles` to plot the remaining peaks on the map. Increase the size of the bubbles based on the elevation of the peaks.

```{r}
tm_shape(mp_above_median) + 
 tm_bubbles(id="name", col =  "ele", size = "ele")
 
```

